

# Ordination and Statistical Analysis

```{r setup-0500-results-ordination}
knitr::opts_chunk$set(fig.path = "fig/0500-results-ordination/", dev = "png")
```


```{r ordination-data}
# Load species matrix
sp_mat_raw <- readr::read_csv("data/processed/benthic_species_matrix.csv",
                               show_col_types = FALSE)

# Parse site and replicate from sample_id
sp_meta <- sp_mat_raw |>
  dplyr::mutate(
    site = stringr::str_extract(sample_id, "^BUL-\\d+"),
    replicate = stringr::str_extract(sample_id, "[A-C]$")
  )

# Community matrix (samples x species, no metadata cols)
sp_mat <- sp_mat_raw |>
  tibble::column_to_rownames("sample_id") |>
  as.matrix()

# Site grouping factor
site_groups <- sp_meta$site
```

## NMDS Ordination

Non-metric multidimensional scaling (NMDS) was used to visualize community similarity among the nine benthic samples (three sites, three replicates each). Ordination was performed on Bray-Curtis dissimilarities of the species abundance matrix using the `vegan` package [@oksanen_etal2024veganCommunity], with two dimensions (k = 2).

<br>

```{r nmds-run}
set.seed(42)
nmds <- vegan::metaMDS(sp_mat, k = 2, distance = "bray", trymax = 999,
                       autotransform = TRUE, trace = 0)

cat("NMDS stress:", round(nmds$stress, 4), "\n")
cat("Converged:", nmds$converged, "\n")
```

<br>

```{r nmds-stress-note, results='asis'}
stress_val <- round(nmds$stress, 3)
stress_interp <- dplyr::case_when(
  stress_val < 0.05 ~ "excellent representation",
  stress_val < 0.10 ~ "good representation",
  stress_val < 0.20 ~ "adequate representation",
  TRUE ~ "poor representation; interpret with caution"
)
cat(paste0(
  "The two-dimensional NMDS solution achieved a stress value of ",
  stress_val, " (", stress_interp,
  "; @clarke_warwick2001Changemarine). "
))
```

Stress measures how well the two-dimensional plot preserves the original dissimilarities among samples — a value below 0.10 means the distances between points on the plot closely reflect the actual differences in community composition. Put simply, samples that appear close together on the ordination really do share similar species, and samples that appear far apart really are compositionally distinct.

<br>

```{r nmds-plot, fig.cap='Two-dimensional NMDS ordination of benthic invertebrate communities across nine samples (three sites, three replicates). Points are coloured by site with spider lines connecting replicates to group centroids and standard deviation ellipses. Closer points indicate more similar community composition.', fig.width=8, fig.height=6}
site_factor <- factor(sp_meta$site, levels = c("BUL-01", "BUL-04", "BUL-05"))
site_colours <- c("BUL-01" = "#1f78b4", "BUL-04" = "#33a02c", "BUL-05" = "#e31a1c")

# Extract scores
scores_sites <- as.data.frame(vegan::scores(nmds, display = "sites"))
scores_sites$site <- sp_meta$site
scores_sites$sample_id <- sp_mat_raw$sample_id

# Compute centroids
centroids <- scores_sites |>
  dplyr::group_by(site) |>
  dplyr::summarise(NMDS1 = mean(NMDS1), NMDS2 = mean(NMDS2), .groups = "drop")

# Spider segments (replicate to centroid)
spider_segs <- scores_sites |>
  dplyr::left_join(centroids, by = "site", suffix = c("", "_c"))

# Plot
p <- ggplot2::ggplot() +
  # Spider lines
  ggplot2::geom_segment(
    data = spider_segs,
    ggplot2::aes(x = NMDS1, y = NMDS2, xend = NMDS1_c, yend = NMDS2_c, colour = site),
    linewidth = 0.4, alpha = 0.5
  ) +
  # SD ellipses
  ggplot2::stat_ellipse(
    data = scores_sites,
    ggplot2::aes(x = NMDS1, y = NMDS2, colour = site),
    type = "norm", level = 0.68, linewidth = 0.8, linetype = "dashed"
  ) +
  # Points
  ggplot2::geom_point(
    data = scores_sites,
    ggplot2::aes(x = NMDS1, y = NMDS2, colour = site, shape = site),
    size = 3.5
  ) +
  # Centroid markers
  ggplot2::geom_point(
    data = centroids,
    ggplot2::aes(x = NMDS1, y = NMDS2, fill = site),
    shape = 23, size = 4, colour = "black", stroke = 0.8
  ) +
  # Sample labels
  ggplot2::geom_text(
    data = scores_sites,
    ggplot2::aes(x = NMDS1, y = NMDS2, label = sample_id),
    size = 2.2, nudge_y = 0.03, alpha = 0.7
  ) +
  ggplot2::scale_colour_manual(values = site_colours) +
  ggplot2::scale_fill_manual(values = site_colours) +
  ggplot2::coord_equal() +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::labs(
    x = "NMDS1", y = "NMDS2",
    caption = paste0("Stress = ", round(nmds$stress, 3))
  ) +
  ggplot2::theme(
    legend.title = ggplot2::element_blank(),
    legend.position = "right"
  )
p
```

<br>

The NMDS ordination revealed clear separation among all three sites, with replicate samples clustering tightly within each site group. BUL-05 replicates formed the tightest cluster, indicating highly consistent community composition across sampling locations at that site. BUL-04 showed the greatest within-site spread, consistent with the high replicate variability observed in univariate metrics (Chironomidae 6–37%, Ephemeroptera 30–63%). BUL-01 replicates clustered together but were well-separated from both BUL-04 and BUL-05, reflecting the distinct Hydropsychidae-dominated community described in the preceding sections.

## PERMANOVA

Permutational multivariate analysis of variance [PERMANOVA; @anderson2001newmethod] was used to test whether benthic community composition differed significantly among the three sites, using the `adonis2` function in `vegan` with 9999 permutations of Bray-Curtis dissimilarities.

<br>

```{r permanova}
set.seed(42)
perm <- vegan::adonis2(sp_mat ~ site_groups, method = "bray", permutations = 9999)
perm
```

<br>

```{r permanova-interp, results='asis'}
f_val  <- round(perm$F[1], 2)
r2_val <- round(perm$R2[1], 3)
p_val  <- perm$`Pr(>F)`[1]

p_text <- if (p_val < 0.001) "p < 0.001" else paste0("p = ", round(p_val, 3))

cat(paste0(
  "Site explained ", round(r2_val * 100, 1), "% of the total variation in community composition ",
  "(F = ", f_val, ", ", p_text, ", 9999 permutations). "
))

if (p_val < 0.05) {
  cat("Community composition differed significantly among sites.")
} else {
  cat("Differences among sites were not statistically significant at the 0.05 level.")
}
```

## Multivariate Dispersion

The PERMANOVA assumption of homogeneous multivariate dispersion was assessed using `betadisper`, which tests whether group centroids differ in their distance to group members (i.e., whether within-site variability is similar across sites).

<br>

```{r betadisper}
bray_dist <- vegan::vegdist(sp_mat, method = "bray")
bd <- vegan::betadisper(bray_dist, site_groups)
set.seed(42)
bd_perm <- permutest(bd, permutations = 9999)
bd_perm
```

<br>

```{r betadisper-interp, results='asis'}
bd_p <- bd_perm$tab$`Pr(>F)`[1]
bd_text <- if (bd_p < 0.001) "p < 0.001" else paste0("p = ", round(bd_p, 3))

if (bd_p >= 0.05) {
  cat(paste0(
    "Multivariate dispersion did not differ significantly among sites (",
    bd_text,
    "), supporting the validity of the PERMANOVA result."
  ))
} else {
  cat(paste0(
    "Multivariate dispersion differed significantly among sites (",
    bd_text,
    "), suggesting that the PERMANOVA result may be partially driven by differences ",
    "in within-site variability rather than shifts in community composition alone."
  ))
}
```

## Pairwise Comparisons

Pairwise PERMANOVA with Holm correction for multiple comparisons was used to identify which specific site pairs differed significantly in community composition.

<br>

```{r pairwise-permanova}
set.seed(42)
pw <- RVAideMemoire::pairwise.perm.manova(
  bray_dist,
  site_groups,
  nperm = 9999,
  p.method = "holm"
)
pw
```

<br>

```{r pairwise-interp, results='asis'}
pw_mat <- pw$p.value
pairs <- which(!is.na(pw_mat), arr.ind = TRUE)

cat("Pairwise PERMANOVA results (Holm-adjusted):\n\n")
for (i in seq_len(nrow(pairs))) {
  r <- pairs[i, "row"]
  c_idx <- pairs[i, "col"]
  p <- pw_mat[r, c_idx]
  p_str <- if (p < 0.001) "p < 0.001" else paste0("p = ", round(p, 3))
  sig <- if (p < 0.05) " *" else ""
  cat(paste0("- ", rownames(pw_mat)[r], " vs. ", colnames(pw_mat)[c_idx],
             ": ", p_str, sig, "\n"))
}
```

<br>

Together, the multivariate analyses confirm that the three Neexdzii Kwa mainstem sites support statistically distinct benthic communities. The tight clustering of replicates within sites demonstrates that the observed differences are not an artefact of sampling variability — each site consistently produces a characteristic assemblage. The high proportion of variance explained by site identity, combined with homogeneous dispersion, provides strong evidence that the spatial gradient in community composition described in the preceding sections reflects genuine ecological differences among reaches.

<!-- ## Indicator Species Analysis

Indicator species analysis will be added when environmental/habitat data from field forms is available, enabling `indicspecies::multipatt()` with site groupings informed by environmental gradients. -->

<!-- ## Environmental Correlation (envfit)

Environmental variable correlation with community structure via `envfit()` will be added when habitat data (substrate, embeddedness, water quality parameters) from field forms is incorporated. -->

